Untargeted control doc

### 2023-11-07 untargeted/paruntargeted_control.Rmd
Restructured isotope detection

# polarity <- "pos"
polarity <- "neg"
# amendment <- "Amm"
# amendment <- "Nit"
# amendment <- "GMP"
amendment <- "Arg"
IS_type <- "noIS"
# IS_type <- "wIS"

@@

mz_group <- function(mz_vals, ppm){
  group_vec <- numeric(length(mz_vals))
  group_num <- 1L
  init_vec <- mz_vals
  names(init_vec) <- seq_along(init_vec)
  while(length(init_vec)>0){
    mz_i <- init_vec[1]
    err <- mz_i*ppm/1000000
    mz_idxs <- init_vec>mz_i-err & init_vec<mz_i+err
    group_vec[as.numeric(names(mz_idxs)[mz_idxs])] <- group_num
    init_vec <- init_vec[!mz_idxs]
    group_num <- group_num+1L
  }
  group_vec
}
trapz <- function(x, y) {
  y <- y[order(x)]
  x <- x[order(x)]
  m <- length(x)
  xp <- c(x, x[m:1])
  yp <- c(numeric(m), y[m:1])
  n <- 2*m
  p1 <- sum(xp[1:(n-1)]*yp[2:n]) + xp[n]*yp[1]
  p2 <- sum(xp[2:n]*yp[1:(n-1)]) + xp[1]*yp[n]

@@

```{r making peak_data and best_feat_data}
msnexp_filled <- readRDS(paste0(output_folder, "msnexp_filled.rds"))
peak_data_long <- msnexp_filled %>%
  chromPeaks() %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  mutate(peakidx=row_number())
peak_data <- msnexp_filled %>%
  featureDefinitions() %>%
  as.data.frame() %>%
  select(mzmed, rtmed, npeaks, peakidx) %>%
  rownames_to_column("id") %>%
  unnest_longer(peakidx) %>%
  rename_with(~paste0("feat_", .x), .cols = -peakidx) %>%
  dplyr::rename(feature="feat_id") %>%
  left_join(peak_data_long, by = join_by(peakidx)) %>%
  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
  select(feature, feat_mzmed, feat_rtmed, feat_npeaks, filename,
         mz, mzmin, mzmax, rt, rtmin, rtmax, into, sn, beta_cor, beta_snr)
output_csv_write(peak_data)

@@

# Rescue feats
if(amendment=="Amm"){
  if(polarity=="pos"){
    rescue_feats <- c("FT3015", Allopurinol="FT1319", Hypoxanthine="FT1321",
                      `DMSOP`="FT1688", "Guan from guan"="FT1707", Tyrosine="FT2200",
                      Phosphocholine="FT2228", Trimethyllysine="FT2278",
                      Dexpanthenol="FT2426", Cytidine="FT2792",
                      Mycogly='FT2814', Inosine="FT2960", Guanosine="FT3015")
  } else if(polarity=="neg"){
    rescue_feats <- c()
  }
}
if(amendment=="Nit"){
  if(polarity=="pos"){
    rescue_feats <- c(Creatinine="FT1339", `Homocysteine thiolactone`="FT1444",
                      Leucine="FT1956", Ornithine="FT1980", Lysine="FT2404",
                      THB="FT2706", Lysinebetaineish='FT3317', glucogly="FT4428")

  } else if(polarity=="neg"){
    rescue_feats <- c(Bisulfate="FT0706", "Itaconic acid"="FT1544", 
                      "Salicylic"="FT1841")
  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="GMP"){
  if(polarity=="pos"){
    rescue_feats <- c("FT1422", "FT1580", "FT1585", "FT1603", "FT3030")
  } else if(polarity=="neg"){
    rescue_feats <- c("FT0865")
  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="Arg"){
  if(polarity=="pos"){
    rescue_feats <- c("FT0945", "FT1138", "FT1346", "FT1413", "FT1476",
                      "FT1505", "FT1533", "FT1935", "FT1945", "FT2029",
                      "FT2425", "FT2774", "FT2999", "FT3668")
  } else if(polarity=="neg"){
    rescue_feats <- c("FT0790", "FT1022", "FT1031", "FT1387")
  } else {
    stop("Unknown polarity")
  }
}

@@

@@

>>>>

polarity <- "pos"
# polarity <- "neg"
# amendment <- "Amm"
# amendment <- "Nit"
amendment <- "GMP"
# amendment <- "Arg"
IS_type <- "noIS"
# IS_type <- "wIS"

@@

mz_group <- function(mz_vals, ppm){
  group_vec <- numeric(length(mz_vals))
  group_num <- 1L
  init_vec <- mz_vals
  names(init_vec) <- seq_along(init_vec)
  while(length(init_vec)>0){
    mz_i <- init_vec[1]
    err <- mz_i*ppm/1000000
    mz_idxs <- init_vec>mz_i-err & init_vec<mz_i+err
    group_vec[as.numeric(names(mz_idxs)[mz_idxs])] <- group_num
    init_vec <- init_vec[!mz_idxs]
    group_num <- group_num+1L
  }
  group_vec
}
trapz <- function(x, y, na.rm=TRUE) {
  y <- y[order(x)]
  x <- x[order(x)]
  m <- length(x)
  xp <- c(x, x[m:1])
  yp <- c(numeric(m), y[m:1])
  n <- 2*m
  p1 <- sum(xp[1:(n-1)]*yp[2:n], na.rm = na.rm) + xp[n]*yp[1]
  p2 <- sum(xp[2:n]*yp[1:(n-1)], na.rm = na.rm) + xp[1]*yp[n]

@@

```{r making peak_data and best_feat_data}
msnexp_filled <- readRDS(paste0(output_folder, "msnexp_filled.rds"))
peak_data_long <- msnexp_filled %>%
  chromPeaks() %>%
  as.data.frame() %>%
  rownames_to_column() %>%
  mutate(peakidx=row_number())
peak_data <- msnexp_filled %>%
  featureDefinitions() %>%
  as.data.frame() %>%
  select(mzmed, rtmed, npeaks, peakidx) %>%
  rownames_to_column("id") %>%
  unnest_longer(peakidx) %>%
  rename_with(~paste0("feat_", .x), .cols = -peakidx) %>%
  dplyr::rename(feature="feat_id") %>%
  left_join(peak_data_long, by = join_by(peakidx)) %>%
  mutate(filename=basename(fileNames(msnexp_filled))[sample]) %>%
  select(feature, feat_mzmed, feat_rtmed, feat_npeaks, filename,
         mz, mzmin, mzmax, rt, rtmin, rtmax, into, sn, beta_cor, beta_snr) %>%
  arrange(desc(into)) %>%
  group_by(feature, filename) %>%
  slice(1) %>%
  ungroup()
output_csv_write(peak_data)

@@

# Rescue feats
if(amendment=="Amm"){
  if(polarity=="pos"){
    rescue_feats <- c("FT3015", Allopurinol="FT1319", Hypoxanthine="FT1321",
                      `DMSOP`="FT1688", "Guan from guan"="FT1707", Tyrosine="FT2200",
                      Phosphocholine="FT2228", Trimethyllysine="FT2278",
                      Dexpanthenol="FT2426", Cytidine="FT2792",
                      Mycogly='FT2814', Inosine="FT2960", Guanosine="FT3015",
                      Hydroxyproline="FT1237")
  } else if(polarity=="neg"){
    rescue_feats <- c()
  }
}
if(amendment=="Nit"){
  if(polarity=="pos"){
    rescue_feats <- c(Creatinine="FT1339", `Homocysteine thiolactone`="FT1444",
                      Leucine="FT1956", Ornithine="FT1980", Lysine="FT2404",
                      THB="FT2706", Lysinebetaineish='FT3317', glucogly="FT4428")

  } else if(polarity=="neg"){
    rescue_feats <- c(Bisulfate="FT0706", "Itaconic acid"="FT1544", 
                      "Salicylic"="FT1841")
  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="GMP"){
  if(polarity=="pos"){
    rescue_feats <- c("FT1422", "FT1580", "FT1585", "FT1603", "FT3030", "FT1569")
  } else if(polarity=="neg"){
    rescue_feats <- c("FT0865")
  } else {
    stop("Unknown polarity")
  }
}
if(amendment=="Arg"){
  if(polarity=="pos"){
    rescue_feats <- c("FT0945", "FT1138", "FT1346", "FT1413", "FT1476",
                      "FT1505", "FT1533", "FT1935", "FT1945", "FT2029",
                      "FT2425", "FT2774", "FT2999", "FT3668")
  } else if(polarity=="neg"){
    rescue_feats <- c("FT0790", "FT1022", "FT1031", "FT1387")
  } else {
    stop("Unknown polarity")
  }
}

@@

```{r find isotope pairs among good peaks}
best_feat_data <- read_csv(paste0(output_folder, "best_feat_data.csv"))
peak_data <- read_csv(paste0(output_folder, "peak_data.csv"))
manual_anno <- readxl::read_excel("untargeted/manual_annotations.xlsx") %>%
  filter(polarity==!!polarity) %>%
  filter(amendment==!!amendment) %>%
  select(feature, compound_name)
iso_mzs <- read_csv(paste0(output_folder, "iso_mzs.csv"))

single_lab_matching <- manual_anno %>%
  left_join(iso_pairs) %>%
  mutate(compound_name=paste(compound_name, "x 15N1")) %>%
  drop_na() %>%
  select(feature=feature_iso, compound_name)
double_lab_matching <- single_lab_matching %>%
  left_join(iso_pairs %>% select(-compound_name)) %>%
  mutate(compound_name=str_replace(compound_name, "x 15N1", "x 15N2")) %>%
  drop_na() %>%
  select(feature=feature_iso, compound_name)
treble_lab_matching <- double_lab_matching %>%
  left_join(iso_pairs %>% select(-compound_name)) %>%
  mutate(compound_name=str_replace(compound_name, "x 15N2", "x 15N3")) %>%
  drop_na() %>%
  select(feature=feature_iso, compound_name)
quad_lab_matching <- treble_lab_matching %>%
  left_join(iso_pairs %>% select(-compound_name)) %>%
  mutate(compound_name=str_replace(compound_name, "x 15N3", "x 15N4")) %>%
  drop_na() %>%
  select(feature=feature_iso, compound_name)
quint_lab_matching <- quad_lab_matching %>%
  left_join(iso_pairs %>% select(-compound_name)) %>%
  mutate(compound_name=str_replace(compound_name, "x 15N4", "x 15N5")) %>%
  drop_na() %>%
  select(feature=feature_iso, compound_name)
iso_anno <- manual_anno %>%
  bind_rows(single_lab_matching, double_lab_matching, treble_lab_matching, 
            quad_lab_matching, quint_lab_matching)

carbon_pairs <- best_feat_data %>%
  select(feature, mzmed) %>%
  pull(mzmed, feature) %>%
  dist() %>%
  as.matrix() %>%
  {which(abs(.-1.003355)<0.001, arr.ind = TRUE)} %>%
  as.data.frame() %>%
  filter(row<col) %>%
  left_join(best_feat_data %>% select(feature, mzmed, rtmed) %>% mutate(row=row_number()),
            by="row") %>%
  left_join(best_feat_data %>% select(feature, rtmed, mzmed) %>% mutate(col=row_number()),
            by="col", suffix=c("", "_iso")) %>%
  filter(abs(rtmed_iso-rtmed)<0.1) %>%
  distinct(feature=feature_iso)

iso_pairs <- best_feat_data %>%
  select(feature, mzmed) %>%
  pull(mzmed, feature) %>%
  dist() %>%
  as.matrix() %>%
  {which(abs(.-0.997035)<0.001, arr.ind = TRUE)} %>%
  as.data.frame() %>%
  filter(row<col) %>%
  left_join(best_feat_data %>% select(feature, mzmed, rtmed) %>% mutate(row=row_number()),
            by="row") %>%
  left_join(best_feat_data %>% select(feature, rtmed, mzmed) %>% mutate(col=row_number()),
            by="col", suffix=c("", "_iso")) %>%
  filter(abs(rtmed_iso-rtmed)<0.1) %>%
  select(-row, -col) %>%
  left_join(iso_anno) %>%
  anti_join(carbon_pairs, by="feature")

output_csv_write(iso_pairs)

# (select_iso <- iso_pairs %>% filter(feature=="FT2466"))
# FT0085_eics <- open_dataset("tmzMLs/pqds") %>%
#   filter(samp_type=="Smp") %>%
#   filter(amendment==!!amendment) %>%
#   filter(polarity==!!polarity) %>%
#   filter(mz%between%pmppm(select_iso$mzmed) | mz%between%pmppm(select_iso$mzmed_iso)) %>%
#   dplyr::collect()
# FT0085_eics %>%
#   filter(rt%between%(select_iso$rtmed+c(-1, 1))) %>%
#   mutate(iso=ifelse(mz<mean(mz), "Unlabeled", "Labeled")) %>%
#   left_join(parametadata) %>%
#   ggplot() +
#   geom_line(aes(x=rt, y=int, group=filename, color=timepoint)) +
#   facet_wrap(~iso, ncol=1, scales="free_y")
```

```{r find isotope envelopes for all untarg}
best_feat_data <- read_csv(paste0(output_folder, "best_feat_data.csv"))
peak_data <- read_csv(paste0(output_folder, "peak_data.csv"))

library(furrr)
plan(multisession, workers = 6)
n_psb_isos <- 6
n_ok_nas <- 10

raw_iso_envelopes <- parametadata %>%
  filter(samp_type=="Smp") %>%
  filter(!str_detect(filename, "Blk")) %>%
  arrange(timepoint, depth, startime) %>%
  pull(filename) %>%
  future_map(function(filename_i){
    file_msdata <- open_dataset("tmzMLs/pqds") %>%
      filter(filename==filename_i) %>%
      dplyr::collect()
    pmap(best_feat_data, function(...){
      row_data <- list(...)
      # row_data <- filter(best_feat_data, feature=="FT0769")
      # row_data <- slice(best_feat_data, 326)
      rt_bounds <- peak_data %>% 
        filter(feature==row_data$feature) %>%
        filter(filename==filename_i) %>%
        reframe(rtmin=rtmin/60, rtmax=rtmax/60)
      if(nrow(rt_bounds)==0)return(NULL)
      file_rt_eic <- file_msdata %>%
        .[rt%between%c(rt_bounds$rtmin, rt_bounds$rtmax)] %>%
        .[mz%between%c(row_data$mzmed-(n_psb_isos+2), row_data$mzmed+(n_psb_isos+2))]

      eic_df <- c(rev(seq(row_data$mzmed, row_data$mzmed-n_psb_isos, -0.997035)[-1]), 
                   seq(row_data$mzmed, row_data$mzmed+n_psb_isos, 0.997035)) %>%
        sapply(pmppm) %>%
        t() %>%
        as.data.frame() %>%
        set_names(c("mzmin", "mzmax")) %>%
        mutate(iso_count=seq(-n_psb_isos, n_psb_isos)) %>%
        left_join(file_rt_eic, by=join_by(mzmin<mz, mzmax>mz)) %>%
        drop_na(rt) %>%
        complete(iso_count, rt) %>%
        group_by(iso_count, rt) %>%
        summarise(int=max(int), .groups = "drop_last") %>%
        filter(sum(!is.na(int))>=n_ok_nas)
      eic_mat <- eic_df %>%
        pivot_wider(names_from = iso_count, values_from = int) %>%
        arrange(rt) %>%
        column_to_rownames("rt")
      if(nrow(eic_mat)==0)return(NULL)
      # scale(eic_mat) %>% matplot(type="l")
      # pairs(eic_mat)
      iso_area_df <- eic_df %>%
        drop_na() %>%
        group_by(iso_count) %>%
        summarise(area=trapz(rt, int))
      eic_mat %>%
        cor(use = "pairwise.complete") %>%
        as.data.frame() %>%
        rownames_to_column("init_count") %>%
        pivot_longer(-init_count, names_to = "iso_count", values_to = "corr") %>%
        mutate(init_count=as.numeric(init_count), iso_count=as.numeric(iso_count)) %>%
        filter(init_count==0) %>%
        select(iso_count, corr) %>%
        mutate(feature=row_data$feature) %>%
        left_join(iso_area_df, by="iso_count")
    }) %>%
      bind_rows() %>% 
      mutate(filename=filename_i) %>%
      select(feature, filename, iso_count, corr, area)
  }, .progress = TRUE) %>%
  bind_rows()
output_csv_write(raw_iso_envelopes)
```

@@

# Export final peak areas
